---
title: "GCRST Approximate Computation"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{GCRST Approximate Computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gammacount)

library(ggplot2)
library(foreach)
library(dplyr)
```

# Background

The Gamma Count with Random Start Time (GCRST) distribution has probability mass function
\begin{align*}
P(X=x) &= \begin{cases}
P(\tau > \lambda), & x=0 \\
\int_0^\lambda P(X'(\tau) = x-1)f_\tau(t) \mathrm{d}t, & x > 0,
\end{cases}
\end{align*}
where $\tau$ is the first event arrival time which follows the first arrival time distribution $f_\tau(t)$, and  $X'(\tau) \sim \mathrm{gc}(\lambda-\tau, \alpha)$. For more details, see the vignette on the derivation of the gamma-count distribution.

For $x=0$, computing this requires only the CDF of the first arrival time distribution, which only depends on `pgamma`. However $x>0$ requires an integral without such a useful form. The default `dgcrst` and `pgcrst` functions use `R`'s built-in `integrate()` function which uses adaptive quadrature to get a fairly accurate answer. But this can be slow, especially because it isn't vectorized.

This vignette is about a method of approximating this integral in a faster way.

# Principle

# Computation Details

## Conditional expected value of $\tau$

## Choice of intervals

# Examples

## Example PMF

```{r exampleplot, fig.width=7, fig.height=5}
lambda <- 4.1
alpha <- 0.7

x <- seq(0, 10)
barplot(
  rbind(dgcrst(x, lambda, alpha),
        dgc(x, lambda, alpha),
        dgcrst.approx(x, lambda, alpha, segments=10)),
  names.arg=x,
  legend.text=c("Quadrature", "No RST", "10 Segment Approx"),
  beside=TRUE
)

```

## Speed comparison

```{r timingcompare, cache=TRUE}
set.seed(2020)

n <- 1000
trials <- 20

lambda <- runif(n, min=0, max=100)
alpha <- rexp(n, rate=log(2))
x <- rpois(n, lambda)

starttime <- Sys.time()
for (i in 1:trials) {
  dgcrst(x, lambda, alpha)
}
stoptime <- Sys.time()

starttime.approx <- Sys.time()
for (i in 1:trials) {
  dgcrst.approx(x, lambda, alpha, segments=10)
}
stoptime.approx <- Sys.time()

starttime.gc <- Sys.time()
for (i in 1:trials) {
  dgc(x, lambda, alpha)
}
stoptime.gc <- Sys.time()

cat(
  "Calculation time with integrate():  ",
  (stoptime - starttime) / trials,
  "\nCalculation time with approximation:",
  (stoptime.approx - starttime.approx) / trials,
  "\nCalculation time assuming no RST:   ",
  (stoptime.gc - starttime.gc) / trials
)

```

## Accuracy measurement

```{r distcompare, cache=TRUE}
lambdas <- c(
  seq(0.5, 20, by=0.5), 
  seq(21, 50, by=1)
)
alphas <- exp(c(
  seq(log(0.1), log(10), length.out=21),
  seq(log(0.01), log(0.1), length.out=5),
  seq(log(10), log(100), length.out=5)
))
approx.levels <- c(1, 5, 10, 50)


differences <- foreach(alpha=alphas, .combine="rbind") %do% {
  foreach(lambda=lambdas, .combine="rbind") %do% {
    # Which x's will we need to compute over?
    x.range <- qgcrst(c(0.0005, 0.9995), lambda, alpha)
    min.x <- x.range[1]
    max.x <- x.range[2]
    x <- seq(min.x, max.x)
    # What is the "true" density and cdf?
    dens.gcrst <- dgcrst(x, lambda, alpha)
    cdf.gcrst <- pgcrst(x, lambda, alpha)
    # Try varying numbers of segments
    approximations <- foreach(approx=approx.levels, .combine="rbind") %do% {
      dens.approx <- dgcrst.approx(x, lambda, alpha, segments=approx)
      cdf.approx <- pgcrst.approx(x, lambda, alpha, segments=approx)
      data.frame(
        alpha=alpha,
        lambda=lambda,
        approximation=paste0("a",approx),
        kl = sum(dens.gcrst * log(dens.gcrst / dens.approx), na.rm=TRUE),
        sumdiff = 0.5 * sum(abs(dens.gcrst - dens.approx)),
        ks = max(abs(cdf.gcrst - cdf.approx)),
        stringsAsFactors=FALSE
      )
    }
    # Use plain GC as an approximation
    dens.gc <- dgc(x, lambda, alpha)
    cdf.gc <- pgc(x, lambda, alpha)
    # Combine the measurements of gc with the rest and return
    rbind(
      approximations,
      data.frame(
        alpha=alpha,
        lambda=lambda,
        approximation="gc",
        kl = sum(dens.gcrst * log(dens.gcrst / dens.gc), na.rm=TRUE),
        sumdiff = 0.5 * sum(abs(dens.gcrst - dens.gc)),
        ks = max(abs(cdf.gcrst - cdf.gc)),
        stringsAsFactors=FALSE
      )
    )
  }
}
```


```{r errorplots, fig.width=7, fig.height=5, }
breaks <- c(0, outer(c(1,5), 10 ^ c(-4, -3, -2, -1)), 1)

differences %>%
  filter(approximation == "gc") %>%
  ggplot(aes(x=lambda, y=alpha, z=sumdiff)) +
  geom_contour_filled(aes(color=sumdiff), breaks=breaks) +
  scale_y_log10()
differences %>%
  filter(approximation == "a1") %>%
  ggplot(aes(x=lambda, y=alpha, z=sumdiff)) +
  geom_contour_filled(aes(color=sumdiff), breaks=breaks) +
  scale_y_log10()
differences %>%
  filter(approximation == "a5") %>%
  ggplot(aes(x=lambda, y=alpha, z=sumdiff)) +
  geom_contour_filled(aes(color=sumdiff), breaks=breaks) +
  scale_y_log10()
differences %>%
  filter(approximation == "a10") %>%
  ggplot(aes(x=lambda, y=alpha, z=sumdiff)) +
  geom_contour_filled(aes(color=sumdiff), breaks=breaks) +
  scale_y_log10()
differences %>%
  filter(approximation == "a50") %>%
  ggplot(aes(x=lambda, y=alpha, z=sumdiff)) +
  geom_contour_filled(aes(color=sumdiff), breaks=breaks) +
  scale_y_log10()
```
